Genetic draft and valley crossing

### 2018-05-13 ms/valley_crossing.tex
re

(Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
Then the fixation probability turns out to be governed by the integro-differential equation
\begin{equation}
v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
\end{equation}
where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.

@@

Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
We have
\begin{align*}
-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
& + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
& +kp(w,T|k-1,t,x).
\end{align*}
Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
We therefore have
\begin{equation}
-\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
\end{equation}
Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
\begin{equation}
\partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
\end{equation}
where $\tilde{\phi}$ = $\phi/\sigma$.
We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
\begin{equation}
\partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
\end{equation}
Using matched asymptotics allows us to write
\begin{equation}
\tilde{\phi} = 
\begin{cases}
\tilde{z} &\mathrm{~if~} \theta < 0, \\
\theta &\mathrm{~if~} \theta > 0.
\end{cases}
\end{equation}
Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.

@@

\begin{figure}
\begin{subfigure}[b]{0.2\textwidth}
\includegraphics[width=\textwidth]{Figures/tau_valley_-6.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\includegraphics[width=\textwidth]{Figures/tau_valley_-2.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\includegraphics[width=\textwidth]{Figures/alpha_-6.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.2\textwidth}
\includegraphics[width=\textwidth]{Figures/alpha_-2.pdf}
\end{subfigure}
\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$. In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$. Left two plots: valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted. Right two plots: the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$. Deeper valleys in smaller populations are disfavored relative to simple sweeps. But at higher values of $\sigma$, they can be favored ($\alpha < 1$: note the log scaled colorbar). In both plots, $N = 10^5$ is a fairly hard cutoff beyond which tunneling becomes semi-deterministic, as $N\mu > 1$ means that, on average, more than one mutant lineage is seeded every generation.}
\label{fig:ratios}
\end{figure}

\begin{figure}
\includegraphics[width=0.4\textwidth]{Figures/alpha_ratio.pdf}
\caption{Ratio of $\alpha$ values at high and low $\sigma$. Genetic draft is most helpful for crossing deep valleys at large population sizes.}
\label{fig:alpha_ratio}
\end{figure}

Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
A higher value of the ratio $\alpha$ means that valley crossing takes much longer than a simple beneficial sweep does.
Though the precise value of this ratio will generally depend on $\delta$, as well as the selection coefficient $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).

>>>>

We begin by considering the dynamics of $p_{\mathrm{fix}}(x,s)$ before moving on to consider the distribution of bubble sizes.
We zero in on the ``weak driver" model, in which background mutations are frequent but of weak effect compared to $\sigma$.
In this case, the background beneficial mutation rate $U_b$ and mean fitness effect $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
Taking the limit $U_b \to \infty$ and $\bar{\epsilon} \to 0$ yields the ``infinitesimal" model.
(note: cite Hallatschek (2013) and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
\begin{equation}
v\partial_x p(x,s) = x p(x,s) + D\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
\end{equation}
in which we have assumed that $v \approx \sigma^2$, which is tantamount to claiming that mutational effects are less important than selection in determining the advance of the mean fitness.
The fixation probability then increases linearly in $x$ above some critical value $x_c$ and remains small otherwise, with $x_c \approx \sigma^4/4D$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).

In order to compute $\Phi$ we must next consider the behavior of mutant lineages.
Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
\citet{neher_shraiman_2011} showed that $\phi$ follows the equation
\begin{equation}
\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - sigma^2 t + \delta)\phi^2,
\end{equation}
but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation or send time to infinity.
Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu p_{\mathrm{fix}}(x)$), means that a simple closed form expression for $\Phi$ is not forthcoming.
We therefore focus on simulation methods to study the dynamics of valley crossing.

\iffalse
(Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
Then the fixation probability turns out to be governed by the integro-differential equation
\begin{equation}
v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
\end{equation}
where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.

@@

Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
We have
\begin{align*}
-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
& + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
& +kp(w,T|k-1,t,x).
\end{align*}
Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
We therefore have
\begin{equation}
-\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
\end{equation}
Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
\begin{equation}
\partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
\end{equation}
where $\tilde{\phi}$ = $\phi/\sigma$.
We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
\begin{equation}
\partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
\end{equation}
Using matched asymptotics allows us to write
\begin{equation}
\tilde{\phi} = 
\begin{cases}
\tilde{z} &\mathrm{~if~} \theta < 0, \\
\theta &\mathrm{~if~} \theta > 0.
\end{cases}
\end{equation}
Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
\fi

@@

\begin{figure}
\begin{subfigure}[b]{0.25\textwidth}
\includegraphics[width=\textwidth]{Figures/tau_valley_-6.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.25\textwidth}
\includegraphics[width=\textwidth]{Figures/tau_valley_-2.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/alpha_ratios.pdf}
\end{subfigure}
\caption{(note: combine left two figures so they have only one colorbar--this will be easy) Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$. In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$. Left two plots: valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted. Right two plots: the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$. Deeper valleys in smaller populations are disfavored relative to simple sweeps. But at higher values of $\sigma$, they can be favored ($\alpha < 1$: note the log scaled colorbar). In both plots, $N = 10^5$ is a fairly hard cutoff beyond which tunneling becomes semi-deterministic, as $N\mu > 1$ means that, on average, more than one mutant lineage is seeded every generation.}
\label{fig:ratios}
\end{figure}

\begin{figure}
\includegraphics[width=0.4\textwidth]{Figures/compare_ratios.pdf}
\caption{Ratio of $\alpha$ values at high and low $\sigma$. Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes. But when semi-deterministic tunneling becomes more important than stochastic tunneling, it can slow down the crossing rate.}
\label{fig:alpha_ratio}
\end{figure}

Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
A higher value of the ratio $\alpha$ means that valley crossing takes much longer than a simple beneficial sweep does.
Though the precise value of this ratio will generally depend on $\delta$, as well as the selection coefficient $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutnats at a low but constant rate.
This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.
